
// Part One //


General Concepts
1. TCGA is a database of cancer patient data. It contains things like patient information, cancer type, 
mutations, RNA expression, methylation, and more. It is important because its a comprehensive collection 
of genomic data, data that is vital for researchers to understand more about different cancers and their 
treatments. 

2. Some strengths of TCGA include its sheer amount of data, as it contains thousands of patients. It has a wide variety of cancers, allowing researchers to study various types. Some of its weaknesses are that it is a public database, meaning it has less funding than private ones. Also, since it is public, there is a lot of data cleaning that needs to be done before any research can be performed.

Coding Skills
1. To save a file to Github from my laptop, I use the following commands:
    - git add         // Prepares file to be added
    - git commit -m   // Allows me to write message
    - git push        // Puts the file into github
    
2. To use a package, you need to download it and then run it.
    - if (!require("package")) install.packages("package")    // Installs package
    - library(package)                                        // Add package to code
    
3. To use a biodconductor package, you need to these steps:
    - if (!require("BiocManager")) install.packages("BiocManager")    // Installs BiocManager
    - if (!require("package")) BiocManager::install("package")        // Installs bioconductor package
    - library(package)                                                // Adds package to code

4. Boolean indexing labels data values as true or false. This is useful because we can then filter out data easily using ifelse mask statements.

5. survival_mask <- data$vital_status == 'Alive'  // This puts TRUE for alive patients and FALSE for dead
   data[survival_mask, ]                          // Filters out rows of patients that are FALSE (dead)
   
   data$age_category <- ifelse(data$age < 30, 'Young',              // Creates new category called age_category
                      ifelse(data$age > 65, 'Old', 'Middle-Aged'))  // that assigns Young, Old, and Middle-Aged
                                                                    // based on the number in data$age
                                                                    

// Part Two //


```{r setup}
    knitr::opts_knit$set(root.dir = normalizePath("/home1/chengpre/490_cluster/analysis_data")) 
```


```{r Package Setup}
# install, only if needed, and load packages 
if (!require("BiocManager")) install.packages("TCGAbiolinks")
if (!require("BiocManager")) install.packages("BiocManager")
if (!require("SummarizedExperiment")) BiocManager::install("SummarizedExperiment")
if (!require("DESeq2")) BiocManager::install("DESeq2")
if (!require("EnhancedVolcano")) BiocManager::install("EnhancedVolcano")
if (!require("maftools")) install.packages("maftools")
if (!require("ggplot2")) install.packages("ggplot2")
if (!require("survival")) install.packages("survival")
if(!require("survminer")) install.packages("survminer")

library(TCGAbiolinks)
library(SummarizedExperiment)
library(DESeq2)
library(EnhancedVolcano)
library(maftools)
library(ggplot2)
library(survival)
library(survminer)
```


















   